---
title: "ACEMS Forecasting Workshop"
author: "Rob J Hyndman"
date: "1.1&nbsp; Introduction to Forecasting"
fontsize: 14pt
output:
  beamer_presentation:
    fig_width: 7
    fig_height: 4.3
    highlight: tango
    theme: metropolis
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache=TRUE,
  warning=FALSE,
  message=FALSE)
library(fpp2)
options(digits=4, width=55)
```

# Background

## Resources

\large

 * Slides
 * Exercises
 * Textbook
 * Useful links

\begin{alertblock}{}{\centerline{\Large\textbf{\url{robjhyndman.com/acemsforecasting2018}}}}
\end{alertblock}

## Key reference

\large

\begin{block}{}\bf
\hangafter=1\hangindent=.3cm
 {Hyndman, R.~J. \& Athanasopoulos, G. (2018) \emph{Forecasting: principles and practice}, 2nd ed.}
\end{block}\pause
\begin{alertblock}{}\Large
\centerline{\bf OTexts.org/fpp2/}
\end{alertblock}

\pause

  * Free and online
  * Data sets in associated R package
  * R code for examples

## Install required packages

```r
install.packages("fpp2", dependencies=TRUE)
```


## Approximate outline
\vspace*{-0.1cm}\centering\fontsize{12}{12}\sf
\begin{tabular}{rp{6.6cm}l}
  \bf Day & \bf Topic                     & \bf Chapter \\
  \midrule
  1.2     & Benchmark methods             & 3 \\
  1.3     & Forecast evaluation           & 3 \\
  2.1     & Transformations               & 3 \\
  2.3     & Stationarity and differencing & 8 \\
  2.4     & Non-seasonal ARIMA models     & 8 \\
  2.5     & Seasonal ARIMA models         & 8 \\
  3.1     & Dynamic regression            & 9 \\
  3.3     & Hierarchical forecasting      & 10 \\
  3.4     & Ensembles                     & 12
\end{tabular}


# The statistical forecasting perspective

## Sample futures

```{r austa1, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE, fig.width=9, fig.height=6}
fit <- ets(austa)
df <- cbind(austa, simulate(fit,10))
for(i in seq(9))
  df <- cbind(df, simulate(fit,10))
colnames(df) <- c("Data", paste("Future",1:10))
autoplot(df) +
  ylim(min(austa),10) +
  ylab("Millions of visitors") + xlab("Year") +
  ggtitle("Total international visitors to Australia") +
 scale_colour_manual(values=c('#000000',rainbow(10)),
                     breaks=c("Data",paste("Future",1:10)),
                     name=" ") +
  ylim(.85,10.0)
```

## Forecast intervals

```{r austa2, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE, fig.width=8.6, fig.height=6}
autoplot(forecast(fit)) +
  ylab("Millions of visitors") + xlab("Year") +
  ggtitle("Forecasts of total international visitors to Australia") +
  ylim(0.85,10.0)
```

